---
title: "Visualizing Data with R: From Basics to Publication-Ready Graphics--Part 1"
author: "Nevada Bioinformatics Center"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_depth: 3
    toc_float: 
      collapsed: false
      smooth_scroll: true
    code_folding: show
    theme: readable
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1. Introduction


### 1.1 Understanding Data Types


### 1.2 Crafting an Effective Data Story


### 1.3 Why R?


### 1.4 Understanding Long vs Wide Data Format

wide data: each row is one sample
long data: each sample may have several rows (reducing the number of columns)

youcan use tidyr to convert wide to long

melt is another function, Cassandra uses it for correlation matrixes


```{r, message=FALSE, warning=FALSE}

# Install libraries if necessary
if (!requireNamespace("tidyr", quietly = TRUE)) {
  install.packages("tidyr")
}

# Example wide dataset
wide_data <- data.frame(
  Group = c("A", "B", "C"),
  Value_1 = c(10, 15, 20),
  Value_2 = c(30, 25, 35)
)

# Convert wide data to long format
library(tidyr)

long_data <- wide_data %>%
  pivot_longer(
    cols = starts_with("Value"),  # Columns to pivot
    names_to = "Variable",        # New column for variable names
    values_to = "Value"           # New column for values
  )

# View the transformed data
print(long_data)

```


```{r, message=FALSE, warning=FALSE}
wide_again <- long_data %>%
  pivot_wider(
    names_from = "Variable",  # Column names
    values_from = "Value"     # Values for those columns
  )

print(wide_again)

```


### 1.5 Introduction to ggplot2

ggplot wants: data, aesthetics, geometryhe


```{r, message=FALSE}
# Install libraries if necessary
if (!requireNamespace("ggplot2", quietly = TRUE)) {
  install.packages("ggplot2")
}

library(ggplot2)

# Get our dataset
data("mtcars")



```

```{r}

# Basic scatter plot of mpg vs wt from the mtcars dataset 
ggplot(data = mtcars, aes(x = wt, y = mpg)) +
  geom_point()

```



```{r, message=FALSE}
ggplot(mtcars, aes(x = wt, y = mpg, color = factor(cyl))) +
  geom_point(size = 3) +
  geom_smooth(method = "lm", se = FALSE) + #regressuion line showing relationshing between x and y
  labs(title = "Scatter Plot of MPG vs. Weight",
       x = "Weight (1000 lbs)",
       y = "Miles per Gallon",
       color = "Cylinders") +
  theme_classic()

```


```{r, message=FALSE}
# Define data, aes, and geom
ggplot(mtcars, aes(x=wt, y=mpg, color = factor(cyl)))+ #use color as factor to see each cylinder as a discrete variable use "factor"
  geom_point()+
  geom_smooth(method = lm)# add linear model regression line with confidence intervals for each color type

# Define data, aes, and geom, make regression line for entire plot
ggplot(mtcars, aes(x=wt, y=mpg))+
  geom_point(aes(color = factor(cyl)))+ #color is attached to geom point so it doesn't affect the line
  geom_smooth(method = lm)# add linear model regression line for ENTIRE DATA SET


```

Add color to aes
Add color to aes as a factor
Add regression line
Add one regression line

### 2.2 Customizing Plots


```{r, message=FALSE}
# Add one black regression line without se
# Define data, aes, and geom, make regression line for entire plot
ggplot(mtcars, aes(x=wt, y=mpg))+
  geom_point(aes(color = factor(cyl)))+ #color is attached to geom point so it doesn't affect the line
  geom_smooth(method = lm, se = FALSE, color = "black", linetype = "dashed")

#when you put AES in the top of the plot script, it affects everything. 

```

Adjust point size and transparency
Change line type to dashed



### Try it for yourself: 

Change the code to increase the regression line thickness.

```{r eval=FALSE}

ggplot(data = mtcars, aes(x = wt, y = mpg)) +
  geom_point(aes(color = factor(cyl)), size = 6, alpha = 0.8) + #alpha = transparency
  geom_smooth(method = "lm", se = FALSE, color = "black", linetype = "dashed", linewidth = .5 )

```




```{r, message=FALSE}

# Change shape of cyl
# Color by horsepower
ggplot(data = mtcars, aes(x = wt, y = mpg)) +
  geom_point(aes(color = (hp), shape = factor(cyl)), size = 6, alpha = 0.8) + #alpha = transparency
  geom_smooth(method = "lm", se = FALSE, color = "black", linetype = "dashed", linewidth = .5 )

#remove "factor" from color of hp becuase hp is a continuous range, keep cyl as a foctor for shape differentiation

```


### Try it for yourself:
 
 
To make adjustments, let's first take a look at the variables we have to work with:
 
```{r}
head(mtcars)
```

The mtcars dataset contains data on 32 car models, with the following columns:

- **mpg:** Miles per gallon (fuel efficiency).

- **cyl:** Number of cylinders in the engine (4, 6, or 8).

- **disp:** Engine displacement (in cubic inches).

- **hp:** Gross horsepower.

- **drat:** Rear axle ratio.

- **wt:** Weight of the car (in 1000s of pounds).

- **qsec:** 1/4 mile time (time in seconds to cover a quarter mile).

- **vs:** Engine shape (0 = V-shaped, 1 = straight).

- **am:** Transmission type (0 = automatic, 1 = manual).

- **gear:** Number of forward gears.

- **carb:** Number of carburetors.
 
 
 Choose two different variables by which to color and shape the points:
 
 Remember that ggplot is built in layers. Everytime you add a "+" you are adding a new layer. 
 You can reorganize your layers by rearranging the lines
 
 
```{r, eval=FALSE}

ggplot(data = mtcars, aes(x = wt, y = mpg)) +
  geom_point(aes(color = disp,              #Fill in 
                 shape = factor(am)),             #Fill in
             size = 5, alpha = 0.8) +
  geom_smooth(method = "lm", se = FALSE, color = "black", linetype = "dashed")



```


## 3. Summary
